## • 21. Quantifying Interactions {.unnumbered #interaction_effect_sizes}


```{r}
#| code-fold: true
#| message: false
#| warning: false
#| code-summary: "Loading and cleaning data"
library(dplyr)
library(readr)
ril_link <- "https://raw.githubusercontent.com/ybrandvain/datasets/refs/heads/master/clarkia_rils.csv"
rils <- readr::read_csv(ril_link) |>
  dplyr::select(ril, prop_hybrid, petal_area_mm, asd_mm, location, petal_color) |>
  na.omit()
```

```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
library(honestlm)
library(dplyr)
library(readr)
library(tibble)
library(stringr)
library(DT)
library(webexercises)
library(ggplot2)
library(tidyr)
library(janitor)
library(forcats)
source("../../_common.R") 
```



::: {.motivation style="background-color: #ffe6f7; padding: 10px; border: 1px solid #ddd; border-radius: 5px;"}


**Motivating scenario:** You ran a model with interactions, estimated parameters, and even evaluated significance. Now what? The results are hard to think about and explain. To understand the biology, I show you how to translate this model into meaningful predictions, estimates, and measures of explained variation.


**Learning goals: By the end of this section, you should be able to:**  


- Use model-based estimates to describe uncertainty in predicted means and slopes.
- Use partial \(R^2\) to summarize how much additional variation is explained by an interaction.
- Write a concise biological interpretation of these results.

:::

---


```{r}
#| echo: false
a <- lm(prop_hybrid ~ petal_color + petal_area_mm + petal_color:petal_area_mm, data = rils)$coef|> round(digits = 4)
```



```{r}
#| echo: false
#| label: fig-interactioncombo
#| fig-height: 4
#| column: page-right
#| fig-width: 12
#| fig-cap: "Observed and model-predicted relationships between petal area and the proportion of hybrid seed for pink- and white-petaled RILs. Panel A shows the data and the fitted model. Panel B extends the fitted lines to a petal area of zero to show what the model parameters are descibing and how they can mislead us."
#| fig-alt: "Two side-by-side plots compare observed and extrapolated interaction-model predictions. Panel A shows individual RILs and separate fitted lines for pink- and white-petaled flowers across the observed range of petal area; the pink line rises strongly while the white line is nearly flat. Panel B shows only the fitted lines extended to petal area zero, where the white line begins above the pink line before the lines cross outside or near the edge of the biologically relevant range."
library(patchwork)
a <- lm(prop_hybrid ~ petal_color + petal_area_mm + petal_color:petal_area_mm, data = rils)$coef|> round(digits = 4)
plot_a<-ggplot(rils, aes(x = petal_area_mm, y = prop_hybrid, color = petal_color)) +
    geom_point()+
  geom_smooth(method = "lm")+
    labs(title = "Visualizing the model & data",
        x = expression("Petal area (" * mm^2 * ")"),
        y = "Prop. hybrid seed",
        color = "Petal color"
    ) 

line_data <- tibble(
    petal_color = c("pink", "white"),
    x = 0,
    xend = 150,
    y = c(
        a["(Intercept)"],
        a["(Intercept)"] + a["petal_colorwhite"]
    ),
    yend = c(a["(Intercept)"] +a["petal_area_mm"] * 150,
        a["(Intercept)"] + a["petal_colorwhite"] +
            (a["petal_area_mm"] + a["petal_colorwhite:petal_area_mm"]) * 150
    )
)

plot_b<-ggplot(line_data) +
    geom_segment(
        aes(
            x = x,
            xend = xend,
            y = y,
            yend = yend,
            color = petal_color
        ), show.legend = FALSE,linewidth=1,
    ) +

    labs(title = "Visualizing the model",
        x = expression("Petal area (" * mm^2 * ")"),
        y = "(predicted)\nProp. hybrid seed",
        color = "Petal color"
    )



plot_c <-ggplot(rils, aes(x = petal_area_mm, y = prop_hybrid, color = petal_color)) +
    geom_point()+
  geom_lm_smooth(method = "lm")+
    labs(title = "No interactions",
        x = expression("Petal area (" * mm^2 * ")"),
        y = "Prop. hybrid seed",
        color = "Petal color"
    ) 

plot_a + plot_b +plot_annotation(tag_levels = "A")+
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


---

We modelled the proportion of hybrid seeds on a RIL as a function of petal area, petal color, and their interaction. If you recall there was some weirdness - the model returned a large and positive coefficient for `petal_colorwhite`, even though we found no region of our data in which the model predicted that white-petaled RILs would have a higher proportion of hybrid seeds than pink-petaled RILs (@fig-interactioncombo A). 

We realized that this was because the coefficient of `0.2373` for `petal_colorwhite` describes the difference between its expected value at the intercept (i.e. `petal_area_mm` = 0, @fig-interactioncombo B) and that of the pink-petled RIls. This prediction with no biological application as no plants have petals of size zero $mm^2$. 




```{r}
#| echo: false
a <- lm(prop_hybrid ~ petal_color + petal_area_mm + petal_color:petal_area_mm, data = rils)$coef|> round(digits = 4)
enframe(a) |> kable()
```

---

### Biological meaning from models

So, if we can't get simple biological meaning from raw model coefficients, how can we make sense of our models? My best advice is to generate meaningful estimates of predicted means + uncertainty in them, for a few illustrative values of the explanatory variables. 

Let's start by doing this ourselves, by simply plugging in petal areas of 45 and 85 $mm^2$ for white- and pink- petaled RILs into the model with interactions. I will also provide predictions from a linear model with no interactions for reference. Click the section below if you want to see this math:  

`r hide("Math here")`

```{r}
#| column: page-right
model_interaction <- lm(prop_hybrid ~ petal_color * petal_area_mm,data = rils)
```

**Interaction model:**


```{r}
#| echo: false
model_interaction
```

Interaction predictions:

```{r}
# Color: pink; petal area:  45 mm^2 
-.2356 + 45 * .007477                         
# Color: white; petal area: 45 mm^2 
-.2356 + 0.237309  + 45 * (.007477-0.006729)  
 # Color: pink; petal area:  85 mm^2 
-.2356 + 85 * .007477       
# Color: white; petal area: 85 mm^2 
-.2356 + 0.237309  + 85 * (.007477-0.006729)
```

**Additive model:**

```{r}
#| column: page-right
model_additive <- lm(prop_hybrid ~ petal_color + petal_area_mm,  data = rils)
```

```{r}
#| echo: false
model_additive 
```
```{r}
# Color: pink; petal area:  45 mm^2 
0.007045 + 45 * 0.003563                          
# Color: white; petal area: 45 mm^2 
0.007045 - 0.180748  + 45 * 0.003563  
# Color: pink; petal area:  85 mm^2 
0.007045 + 85 * 0.003563                        
# Color: white; petal area: 85 mm^2 
0.007045 - 0.180748  + 85 * 0.003563 
```

`r unhide()`


```{r}
#| echo: false
prediction_data <- tidyr::expand_grid(
  petal_area_mm = c(45, 85),
  petal_color = c("pink", "white")
)

prediction_data$interaction_model <- predict(
  model_interaction,
  newdata = prediction_data
)

prediction_data$additive_model <- predict(
  model_additive,
  newdata = prediction_data
)

prediction_data |>
  mutate(
    across(
      c(interaction_model, additive_model),
      \(x) round(x, 3)
    )
  ) |>
  knitr::kable(
    col.names = c(
      "Petal area",
      "Petal color",
      "Interaction model",
      "No-interaction model"
    ),
    align = "cccc"
  )

```



```{r}
#| echo: false
#| label: fig-interactioncombono
#| fig-height: 4
#| column: page-right
#| fig-width: 12
#| fig-cap: "Predicted relationships between petal area and the proportion of hybrid seed under models with and without an interaction. Panel A shows the interaction model. Panel B shows  the additive model."
#| fig-alt: "Two side-by-side plots compare fitted relationships between petal area and proportion of hybrid seed for pink- and white-petaled RILs. In panel A, the interaction model shows a steep positive slope for pink-petaled RILs and a nearly flat slope for white-petaled RILs, so the distance between the lines increases with petal area. In panel B, the additive model shows parallel lines with the same positive slope, so the difference between petal colors remains constant across petal area."
library(patchwork)

plot_a + plot_c +plot_annotation(tag_levels = "A")+
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


---

In the interaction model, the predicted proportion of hybrid seed increases strongly with petal area among pink-petaled RILs but changes little among white-petaled RILs (@fig-interactioncombono A). So the predicted difference between colors is much larger at $85 mm^2$ than at 45 $mm^2$. Because the additive model predicts the same increase with petal area for both petal colors, the predicted difference between pink- and white-petaled RILs is constant across petal areas (@fig-interactioncombono B).  

### How much does the interaction explain?  


$R^2$, the proportion of variation explained is a standard summary of "effect size." In the section on multiple regression, we saw that we could calculate partial $R^2$ as:  

$$R^2_\text{partial} = 1-\frac{SS_\text{error (full model)}}{SS_\text{error (model without focal predictor)}}$$ 

**For an interaction model we do the same thing, first for a main effects model:**

```{r}
library(honestlm)
lm(prop_hybrid ~ petal_color + petal_area_mm,  data = rils)|>
    partial_r2()
```

We see that petal color and petal area "explain" 16% and 6% of the variance in the proportion of hybrid seeds on a RIL, respectively, after adjusting for the other variable. 

---

**And then for the interaction:**


```{r}
library(honestlm)
lm(prop_hybrid ~ petal_color * petal_area_mm,  data = rils)|>
    partial_r2()
```


--- 

After adjusting for these main effects, the interaction between petal area and petal color explains 5% of the remaining variance in the proportion of hybrid seeds. Thus this interaction explains nearly as much variation  in the proportion of hybrid seeds as petal area itself!!!

### Describing uncertainty   

Of course, no quick statistical descriptions is complete without a measure of uncertainty. Again we can accomplish this with the [`emmeans`](https://rvlenth.github.io/emmeans/) or `modelbased` packages. Here I use [`estimate_means()`](https://easystats.github.io/modelbased/reference/estimate_means.html) function in the [`modelbased`](https://easystats.github.io/modelbased/) package, because I find it easier to code. 


So to find uncertainty about the point estimates of the proportion of hybrid seeds for pink- and white-petaled RILs with petal aread of 45 and 85 $mm^2$, we type: 

```{r}
#| eval: false
library(modelbased)
estimate_means(
    lm(prop_hybrid ~ petal_color * petal_area_mm, data = rils),
    by = list(
        petal_color = c("pink", "white"),
        petal_area_mm = c(45, 85)))
```

```{r}
#| echo: false
library(modelbased)
estimate_means(
    lm(prop_hybrid ~ petal_color * petal_area_mm, data= rils),
    by = list(
        petal_color = c("pink", "white"),
        petal_area_mm = c(45,85) ))|>
  kable(digits=3)
```

Here, each row is a model-predicted mean at one color–area combination, the confidence interval describes uncertainty around that predicted mean. 

---

We can also report uncertainty in the slope for each group with the [`estimate_slopes()`](https://easystats.github.io/modelbased/reference/estimate_slopes.html) function in the  [`modelbased`](https://easystats.github.io/modelbased/) package:

```{r}
#| eval: false
library(modelbased)
estimate_slopes(
    lm(prop_hybrid ~ petal_color * petal_area_mm, data= rils),
    slope = "petal_area_mm",    
    by = list(petal_color = c("pink", "white")))
```

```{r}
#| echo: false
estimate_slopes(
    lm(prop_hybrid ~ petal_color * petal_area_mm, data= rils),
    slope = "petal_area_mm",
    by = list( petal_color = c("pink", "white") ))|>
  kable(digits=4)
```

The two confidence intervals describe uncertainty in each slope. Note that they  do not directly test the difference between slopes, as we do this by testing the interaction. 

---


### Writing up our results  

Let's try to synthesize these results into a brief "results" paragraph: 

*We tested the contribution of petal color, petal area and their interaction to the proportion of hybrid seeds on a RIL. Using Type II sums of squares, we found significant associations between the proportion of hybrid seeds and petal color ($F_{1,398} = 80.2$, p < 2.2e-16), petal area ($F_{1,398} = 28.1$, p < 2 e-7), and their interaction ($F_{1,398} = 24.5$, p < 2 e-6). This interaction reflects the sharp increase in the proportion of hybrid seed with petal area in pink-, but not white-petaled RILs (compare slopes of 0.0075 -- 95% CI: {0.0054 : 0.0095}  and 0.0007 -- 95% CI: {-0.0010 : 0.0025}, respectively).*



::: aside
.
:::