## • 9. Bootstrap script {.unnumbered #boot_script}


```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
```

## Overview

Here I provide streamlined versions of R scripts to estimate uncertainty by a bootstrap.


```{r}
#| code-fold: true
#| message: false
#| warning: false
#| code-summary: "Code for selecting data from a few columns from RILs planted at GC"
library(readr)
library(dplyr)
ril_link <- "https://raw.githubusercontent.com/ybrandvain/datasets/refs/heads/master/clarkia_rils.csv"
ril_data <- readr::read_csv(ril_link) 

gc_rils <- ril_data |>
  filter(location == "GC", !is.na(prop_hybrid), ! is.na(mean_visits))|>
  select(petal_color, petal_area_mm, num_hybrid, offspring_genotyped, prop_hybrid, mean_visits , asd_mm )|>
  mutate(id = factor(1:n()))
```

### Functions

The key "workhorse" functions I will use are: 

- The [`specify()`](https://infer.tidymodels.org/reference/specify.html) to [`generate()`](https://infer.tidymodels.org/reference/generate.html) to [`calculate()`](https://infer.tidymodels.org/reference/calculate.html) pipeline from the [`infer`](https://infer.tidymodels.org/) package.  

- [`summarize()`](https://dplyr.tidyverse.org/reference/summarise.html) from [`dplyr`](https://dplyr.tidyverse.org/).    

- [`sd()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/sd.html) from base R. We'll use this to get the standard deviation of the bootstrap distribution - aka the "bootstrap standard error."    

- [`quantile()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/quantile.html) from base R. We'll use this to get the upper and lower tails of the bootstrap distirbution to find the  "bootstrap confidence interval." 

### Reproducible script   

Now here it is!


#### Set up

```{r}
#| message: false
#| warning: false
library(dplyr)
library(infer)
alpha   <- .05
n_boots <- 5000
```

--- 

#### For mean prop_hybrid


```{r}
#| echo: false
#| column: margin

 tibble(se = 0.023, lower_ci = 0.107, upper_ci = 0.197)|>kable()
```

```{r}
#| message: false
#| warning: false
#| eval: false
library(dplyr)
library(infer)
gc_rils %>%
    specify(response = prop_hybrid)               |> 
    generate(reps = n_boots, type = "bootstrap")  |> 
    calculate(stat = "mean")|>
    summarise(se = sd(stat),
            lower_ci = quantile(stat, probs = alpha /2,),
            upper_ci = quantile(stat, probs = 1 - alpha /2))
```


--- 

#### For differences in mean prop_hybrid by petal color


```{r}
#| echo: false
#| column: margin
 tibble(se = 0.042, lower_ci = 0.149, upper_ci = 0.312)|>kable()
```

```{r}
gc_rils                                                          |>
  filter(!is.na(petal_color))                                    |>
  specify(prop_hybrid ~ petal_color)                             |>                      
  generate(reps = n_boots, type = "bootstrap")                   |>                   
  calculate(stat = "diff in means", order = c("pink", "white"))  |>
  summarise(se = sd(stat),
            lower_ci = quantile(stat, probs = alpha /2,),
            upper_ci = quantile(stat, probs = 1 - alpha /2))
```

---

#### For differences in mean prop_hybrid by petal color


```{r}
#| echo: false
#| column: margin
 tibble(se = 0.042, lower_ci = 0.149, upper_ci = 0.312)|>kable()
```

```{r}
gc_rils                                                          |>
  filter(!is.na(petal_color))                                    |>
  specify(prop_hybrid ~ petal_color)                             |>                      
  generate(reps = n_boots, type = "bootstrap")                   |>                   
  calculate(stat = "diff in means", order = c("pink", "white"))  |>
  summarise(se = sd(stat),
            lower_ci = quantile(stat, probs = alpha /2,),
            upper_ci = quantile(stat, probs = 1 - alpha /2))
```

---



#### For slopes


```{r}
#| echo: false
#| column: margin
 tibble(se = 0.00173, lower_ci = 0.00217, upper_ci = 0.00896)|>kable()
```

```{r}
gc_rils                                           |>
  filter(!is.na(petal_area_mm))                   |>
  specify(prop_hybrid ~ petal_area_mm)            |>  # The linear model we're looking into
  generate(reps = n_boots, type = "bootstrap")    |>  # Bootstrapping
  calculate(stat = "slope")                       |>
  summarise(se = sd(stat),
            lower_ci = quantile(stat, probs = alpha /2,),
            upper_ci = quantile(stat, probs = 1 - alpha /2))
```
