## • 16. Two-sample t-test {.unnumbered #nhs2t}

```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
library(dplyr)
library(readr)
library(stringr)
library(DT)
library(webexercises)
library(ggplot2)
library(tidyr)
library(gt)
source("../../_common.R") 
```

::: {.motivation style="background-color: #ffe6f7; padding: 10px; border: 1px solid #ddd; border-radius: 5px;"}

**Motivating scenario:** We want to test the null that two groups have the same mean.

**Learning goals:** By the end of this section, you should be able to:   

-  **Explain the logic of the two-sample t-test** and how it relates to comparing group means.

- **Calculate the test statistic and p-value** by hand and interpret it relative to a critical value.  

- **Use R’s t.test() function** to test for differences between two groups.  

- **Write up results of a two-sample t-test** in terms of both statistical significance and effect size, and connect them back to biological meaning.

:::


```{r}
#| code-fold: true
#| code-summary: "Loading and processing data"
ril_link <- "https://raw.githubusercontent.com/ybrandvain/datasets/refs/heads/master/clarkia_rils.csv"
SR_visits <- readr::read_csv(ril_link) |>
    dplyr::rename(visits = mean_visits)  |>
    dplyr::filter(location == "SR",
                  !is.na(petal_color),
                  !is.na(visits))        |>
    dplyr::select(petal_color, visits)

color_visit_summaries <- SR_visits |>
  group_by(petal_color)|>
  summarise(MEAN  = mean(visits),
            VAR   = var(visits),
            N     = n())
```


## Testing the null of equal means  

In the previous section, we calculated a difference between pink- and white- petaled RILs at site SR of $1.76 - 0.733 = 1.027$.  We also quantified uncertainty about this estimate as a standard error of 0.262. Recalling the equation for the t-statistic:

$$t = \frac{\bar{x}-\mu_0}{s_\bar{x}}  = \frac{(\bar{x_1}-\bar{x_2})-0}{s_\overline{x_1-x_2}}$$ 
In this case 

$$t = \frac{1.76 - 0.733}{0.262}= 3.91$$ 



Since we had  105 degrees of freedom, our critical t-value will be just a hair greater than 1.96. Because our observed t-value of $\approx 4$ is way bigger than our critical t-value of $\approx 2$, we strongly reject the null hypothesis that pink and white-flowered *Clarkia xantiana subspecies parviflora* RILs at site SR are visited equally by pollinators.  

:::aside
More precisely, *crit-t* = `qt(0.025, df= 105,lower.tail = F)` = $1.983$.
:::

We can use the `pt()` function to quantify exactly how rarely the null would generate a result this or more extreme as follows:

`2 * pt(q = 3.91, df = 105, lower.tail = FALSE)` =  `0.000164 = 1.64e-4` 

:::exercises
**Review:** Why do we multiply our p-value by two? `r mcq(c("Because p-values from t-tests are biased to be too low",answer = "To capture both ways in which the data could deviate from the null","We shouldn't. Yaniv messed up"))`
:::

## A two-sample t-test in R 

We can use the "formula" syntax in the [`t.test()`](https://stat.ethz.ch/R-manual/R-patched/library/stats/html/t.test.html) function to have R test this null for us. Note that we set `var.equal = TRUE`. We see that this provides p-values and 95% confidence intervals identical to what we calculated ourselves:    

```{r}
t.test(visits ~ petal_color, data = SR_visits, var.equal = TRUE)
```

Again, we can make this output tidy / easier to process in R with the [`tidy()`](https://generics.r-lib.org/reference/tidy.html) function in the [`broom` package](https://broom.tidymodels.org/): 

:::aside 
For some reason  that I don't understand, [`tidy()`](https://generics.r-lib.org/reference/tidy.html) labels the column with the "degrees of freedom" as "parameter". smdh.
:::

```{r}
#| eval: false
library(broom)
t.test(visits ~ petal_color, data = SR_visits, var.equal = TRUE)|>
  tidy()
```

```{r}
#| echo: false
#| column: page-right

library(broom)
t.test(visits ~ petal_color, data = SR_visits, var.equal = TRUE)|>
  tidy() |>
  mutate_at(c(1:4,7:8) ,round, digits = 3)|>
  mutate(p.value= c("1.65e-4"))|>
  gt()
```



**If our data met assumptions we would be done.**  But they do not. So let's validate our conclusions with tests whose assumptions better match our data.  

### A two-sample t-test is often robust  

Now that we know we can reject the null when data are transformed to meet assumptions of the two-sample t-test, and we have an estimate of the bootstrap 95% confidence interval,  we could complement these analyses with an analysis of the untransformed data. This step is not always necessary or reliable - but we are using our brains  to tell a coherent story. 

**First, let's calculate our summary statistics:**

```{r}
#| echo: false
#| column: margin
SR_visits |>
  group_by(petal_color)|>
  summarise(MEAN = mean(visits),
            VAR  = var(visits),
            N                 = n())|>
  mutate_at(-1, round, digits = 3)|>
  mutate(DF = N-1)|>
  summarise(mean_diff  = diff(MEAN) |> abs(), 
            pooled_var = sum(DF*(VAR)) / (sum(DF)),
            pooled_sd  = sqrt(pooled_var),
            cohens_D   = mean_diff / pooled_sd)|>
    pivot_longer(values_to = "estimate", cols=1:4,names_to = "summary")|>
    mutate(estimate = round(estimate, digits=3))|>
  gt()
```



```{r}
color_visit_summaries <- SR_visits |>
  group_by(petal_color)|>
  summarise(MEAN = mean(visits),
            VAR  = var(visits),
            N                 = n()) |>
    summarise(mean_diff  = diff(MEAN) |> abs(), 
              pooled_var = sum((N-1)*(VAR)) / (sum(N)-2),
              pooled_sd  = sqrt(pooled_var),
              cohens_D   = mean_diff / pooled_sd)
```

**Now let's use R to test the null hypothesis and estimate confidence intervals.** For fun let's compare results from an analysis assuming equal variance to one that does not: 

```{r}
#| eval: false
#| column: page-right
library(broom)

bind_rows(
  t.test(visits ~ petal_color, data = SR_visits, var.equal = TRUE)|> tidy(),
  t.test(visits ~ petal_color, data = SR_visits, var.equal = FALSE)|> tidy()
)
```


```{r}
#| echo: false
#| column: page-right
bind_rows(
  t.test(visits ~ petal_color, data = SR_visits, var.equal = TRUE)|> tidy(),
  t.test(visits ~ petal_color, data = SR_visits, var.equal = FALSE)|> tidy()
)|>
  mutate_at(c(1:4,7:8) ,round, digits = 3)|>
  mutate_at(6 ,round, digits = 1)|>
  mutate(p.value= c("1.65e-4","1.10e-4"))|>
  gt()
```


:::fyi
The [Welch's two-sample t-test](https://en.wikipedia.org/wiki/Welch%27s_t-test) does not assume equal variance, and in practice is universally better than the standard two-sample t-test (that's why R has it as a default). 

$$t = \frac{\overline{X}_1 - \overline{X}_2}{\sqrt{\frac{s_1^2}{N_1} + \frac{s_2^2}{N_2}}}
\text{, and degrees of freedom: } \text{df} \approx \frac{\left(\frac{s_1^2}{N_1} + \frac{s_2^2}{N_2}\right)^2}{\frac{s_1^4}{N_1^2 \times df_1} + \frac{s_2^4}{N_2^2 \times df_2}}$$

However the standard two-sample t-test is often good enough. It also has the benefit of being much like all linear modeling efforts, and is  simpler mathematically, so we usually teach the standard two-sample t-test. In practice the difference between these tests rarely matters, except for when the variance between groups is massively different
:::

### Permutation test 

Of course, rather than using a parametric test, we could find a p-value by a permutation test. As I [introduced previously](#shuffling), permutation tests build a null sampling distribution by shuffling the relation between explanatory and response variables, and comparing our actual difference in means to this permuted distribution. One way to do this in R is with the [pipeline in the `infer` package](https://infer.tidymodels.org/articles/observed_stat_examples.html#one-numerical-variable-one-categorical-2-levels-diff-in-means):


```{r}
#| eval: false
library(infer)
raw_diff <- SR_visits %>%
  specify(visits ~ petal_color) |>              # response ~ explanatory
  calculate(stat = "diff in means",             # difference in group means
            order = c("pink", "white"))

# PERMUTE
perm_SR_visits <-  SR_visits %>%
  specify(visits ~ petal_color) |>              # specify response ~ explanatory
  hypothesize(null = "independence") |>
  generate(reps = 1000000, type = "permute") |>
  calculate(stat = "diff in means", order = c("pink", "white"))

perm_SR_visits  |>
  get_p_value(obs_stat = pull(raw_diff) , direction = "two-sided")
```

```{r}
tibble(p_value = 0.00006)
```


Here, our p-value of 0.00006 is a bit smaller than that estimated from our t-test. But this would not meaningfully change any of our conclusions. 

## Writing up results

Now we can write up our results. Note this takes some thinking because we had to make some decisions. Here's my attempt. 

:::aside
Note that I chose to report values from Welch's t-test (which does not assume equal variance) on untransformed data. To me, this strikes a balance between interpretability and statistical correctness. Although the result is essentially the same as above, the conservatism of Welch's t-test may help build readers' trust. 
:::

*At the Sawmill Road site, pink-flowered Clarkia xantiana ssp. parviflora RILs received, on average, more pollinators during a 15-minute observation than white-flowered RILs (mean pink = 1.76, mean white = 0.73). The mean difference of 1.03 visits was statistically significant (Welch's t = 4.05, df = 89.6, p = 0.00011) and associated with a moderate-to-large effect size (Cohen’s D = 0.75). Results were robust to the right-skew in the data -- bootstrap confidence intervals closely matched analytic ones, and a permutation-based p-value was similarly much smaller than 0.0001. We reject the null and conclude that pink-flowered plants attract more pollinators than white-flowered plants at this site.*

