## • 19. Multiple Regression in R {.unnumbered #multiple_regression_R_workflow}


---
format: html
webr:
  packages: ['dplyr', 'readr' ,'ggplot2']
  autoload-packages: true
---

```{webr-r}
#| context: setup

honest_factor_info <- function(model) {
  mf <- stats::model.frame(model)
  if (ncol(mf) == 0) {
    return(list())
  }

  response <- as.character(stats::formula(model)[[2]])
  predictors <- setdiff(names(mf), response)

  factor_vars <- predictors[vapply(
    mf[predictors],
    function(x) is.factor(x) || is.character(x),
    logical(1)
  )]

  stats::setNames(
    lapply(factor_vars, function(var) {
      x <- mf[[var]]
      if (!is.factor(x)) {
        x <- factor(x)
      }

      list(
        name = var,
        n_levels = nlevels(x),
        levels = levels(x),
        reference = levels(x)[1]
      )
    }),
    factor_vars
  )
}

honest_multilevel_factor_rows <- function(model, terms) {
  factor_info <- honest_factor_info(model)
  rows <- rep(FALSE, length(terms))

  if (length(factor_info) == 0) {
    return(rows)
  }

  for (info in factor_info) {
    if (info$n_levels <= 2) {
      next
    }

    rows <- rows | (startsWith(terms, info$name) & terms != info$name)
  }

  rows
}

honest_contrast_notes <- function(model, terms) {
  factor_info <- honest_factor_info(model)
  notes <- rep(NA_character_, length(terms))

  if (length(factor_info) == 0) {
    return(notes)
  }

  for (info in factor_info) {
    term_matches <- startsWith(terms, info$name) & terms != info$name
    if (!any(term_matches)) {
      next
    }

    if (info$n_levels > 2) {
      notes[term_matches] <- paste0(
        "comparison to reference level `",
        info$reference,
        "`; not a test of whether `",
        info$name,
        "` matters overall"
      )
    } else {
      notes[term_matches] <- paste0(
        "comparison to reference level `",
        info$reference,
        "`"
      )
    }
  }

  notes
}

honest_plot_group_var <- function(data) {
  candidates <- intersect(
    c("colour", "color", "fill", "linetype", "shape"),
    names(data)
  )

  candidates <- candidates[vapply(
    data[candidates],
    function(x) length(unique(x[!is.na(x)])) > 1,
    logical(1)
  )]

  if (length(candidates) == 0) {
    return(NULL)
  }

  candidates[1]
}

format_number <- function(x, digits) {
  format(signif(x, digits), trim = TRUE)
}

`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}


StatLmSmooth <- ggplot2::ggproto(
  "StatLmSmooth",
  ggplot2::Stat,

  required_aes = c("x", "y"),
  extra_params = c("na.rm", "se", "level", "n", "fullrange"),

  compute_layer = function(self, data, params, layout) {
    se <- params$se %||% TRUE
    level <- params$level %||% 0.95
    n <- params$n %||% 100
    fullrange <- params$fullrange %||% FALSE

    keep <- stats::complete.cases(data[, c("x", "y", "group")])
    data <- data[keep, , drop = FALSE]

    if (nrow(data) == 0) {
      return(data[0, , drop = FALSE])
    }

    has_groups <- "group" %in% names(data) && length(unique(data$group)) >= 2

    if (!has_groups) {
      fit <- stats::lm(y ~ x, data = data)

      aesthetic_cols <- intersect(
        c("colour", "color", "fill", "alpha", "linetype", "linewidth", "size"),
        names(data)
      )
      x_global <- range(data$x, na.rm = TRUE)
      panels <- unique(data$PANEL)

      grid <- lapply(panels, function(this_panel) {
        this_data <- data[data$PANEL == this_panel, , drop = FALSE]
        x_range <- if (isTRUE(fullrange)) x_global else range(this_data$x, na.rm = TRUE)

        pred_data <- data.frame(
          x = seq(x_range[1], x_range[2], length.out = n),
          PANEL = this_panel,
          group = -1
        )

        for (aesthetic in aesthetic_cols) {
          pred_data[[aesthetic]] <- this_data[[aesthetic]][1]
        }

        pred_data
      })

      pred_data <- do.call(rbind, grid)
      pred <- stats::predict(fit, newdata = pred_data, se.fit = TRUE)
      pred_data$y <- as.numeric(pred$fit)

      if (isTRUE(se)) {
        crit <- stats::qt((1 + level) / 2, df = stats::df.residual(fit))
        pred_data$se <- as.numeric(pred$se.fit)
        pred_data$ymin <- pred_data$y - crit * pred_data$se
        pred_data$ymax <- pred_data$y + crit * pred_data$se
      }

      return(pred_data)
    }

    data$.lm_group <- factor(data$group)
    fit <- stats::lm(y ~ x + .lm_group, data = data)

    aesthetic_cols <- intersect(
      c("colour", "color", "fill", "alpha", "linetype", "linewidth", "size"),
      names(data)
    )
    x_global <- range(data$x, na.rm = TRUE)
    panel_group_combos <- unique(data[, c("PANEL", "group"), drop = FALSE])

    grid <- lapply(seq_len(nrow(panel_group_combos)), function(i) {
      this_panel <- panel_group_combos$PANEL[i]
      this_group <- panel_group_combos$group[i]
      this_data <- data[data$PANEL == this_panel & data$group == this_group, , drop = FALSE]
      x_range <- if (isTRUE(fullrange)) x_global else range(this_data$x, na.rm = TRUE)

      pred_data <- data.frame(
        x = seq(x_range[1], x_range[2], length.out = n),
        PANEL = this_panel,
        group = this_group
      )
      pred_data$.lm_group <- factor(this_group, levels = levels(data$.lm_group))

      for (aesthetic in aesthetic_cols) {
        pred_data[[aesthetic]] <- this_data[[aesthetic]][1]
      }

      pred_data
    })

    pred_data <- do.call(rbind, grid)
    pred <- stats::predict(fit, newdata = pred_data, se.fit = TRUE)
    pred_data$y <- as.numeric(pred$fit)

    if (isTRUE(se)) {
      crit <- stats::qt((1 + level) / 2, df = stats::df.residual(fit))
      pred_data$se <- as.numeric(pred$se.fit)
      pred_data$ymin <- pred_data$y - crit * pred_data$se
      pred_data$ymax <- pred_data$y + crit * pred_data$se
    }

    pred_data
  }
)
geom_lm_smooth <- function (mapping = NULL, data = NULL, ..., interaction = FALSE, 
    method = NULL, se = TRUE, level = 0.95, n = 100, fullrange = FALSE, 
    na.rm = FALSE, show.legend = NA, inherit.aes = TRUE) 
{
    if (!is.null(method) && !identical(method, "lm")) {
        warning("`geom_lm_smooth()` always uses linear models; ignoring `method = ", 
            deparse(method), "`.", call. = FALSE)
    }
    if (isTRUE(interaction)) {
        return(ggplot2::geom_smooth(mapping = mapping, data = data, 
            method = "lm", formula = y ~ x, se = se, level = level, 
            n = n, fullrange = fullrange, na.rm = na.rm, show.legend = show.legend, 
            inherit.aes = inherit.aes, ...))
    }
    ggplot2::layer(data = data, mapping = mapping, stat = StatLmSmooth, 
        geom = "smooth", position = "identity", show.legend = show.legend, 
        inherit.aes = inherit.aes, params = list(se = se, level = level, 
            n = n, fullrange = fullrange, na.rm = na.rm, ...))
}


StatLmMeans <- ggplot2::ggproto(
  "StatLmMeans",
  ggplot2::Stat,

  required_aes = c("x", "y"),
  extra_params = c("na.rm", "interaction", "se", "level"),

  compute_layer = function(self, data, params, layout) {
    interaction <- params$interaction %||% FALSE
    se <- params$se %||% TRUE
    level <- params$level %||% 0.95

    group_var <- honest_plot_group_var(data)
    needed <- c("x", "y", "PANEL", group_var)
    keep <- stats::complete.cases(data[, needed, drop = FALSE])
    data <- data[keep, , drop = FALSE]

    if (nrow(data) == 0) {
      return(data[0, , drop = FALSE])
    }

    same_partition <- function(a, b) {
      a <- factor(a)
      b <- factor(b)
      nlevels(base::interaction(a, b, drop = TRUE)) == max(nlevels(a), nlevels(b))
    }

    x_levels <- sort(unique(data$x))
    data$.lm_x <- factor(data$x, levels = x_levels)
    x_has_multiple_levels <- length(x_levels) > 1

    panel_levels <- unique(data$PANEL)
    data$.lm_panel <- factor(data$PANEL, levels = panel_levels)
    panel_has_multiple_levels <- nlevels(data$.lm_panel) > 1

    group_has_multiple_levels <- FALSE
    include_group <- FALSE
    if (!is.null(group_var)) {
      data$.lm_group <- factor(data[[group_var]], levels = unique(data[[group_var]]))
      group_has_multiple_levels <- nlevels(data$.lm_group) > 1
      include_group <- group_has_multiple_levels &&
        !same_partition(data$.lm_group, data$.lm_x) &&
        !(panel_has_multiple_levels && same_partition(data$.lm_group, data$.lm_panel))
    }

    include_panel <- panel_has_multiple_levels &&
      !same_partition(data$.lm_panel, data$.lm_x) &&
      !(include_group && same_partition(data$.lm_panel, data$.lm_group))

    terms <- character(0)
    if (x_has_multiple_levels) {
      terms <- c(terms, ".lm_x")
    }
    if (include_group) {
      terms <- c(terms, ".lm_group")
    }
    if (include_panel) {
      terms <- c(terms, ".lm_panel")
    }

    formula <- if (length(terms) == 0) {
      y ~ 1
    } else if (isTRUE(interaction) && include_group && x_has_multiple_levels) {
      stats::as.formula(paste("y ~", paste(c(".lm_x * .lm_group", setdiff(terms, c(".lm_x", ".lm_group"))), collapse = " + ")))
    } else {
      stats::as.formula(paste("y ~", paste(terms, collapse = " + ")))
    }
    fit <- stats::lm(formula, data = data)

    combo_cols <- c("PANEL", "x", group_var)
    combos <- unique(data[, combo_cols, drop = FALSE])
    combos <- combos[order(combos$PANEL, combos$x), , drop = FALSE]

    pred_data <- combos
    pred_data$.lm_x <- factor(pred_data$x, levels = levels(data$.lm_x))
    pred_data$.lm_panel <- factor(pred_data$PANEL, levels = levels(data$.lm_panel))
    if (!is.null(group_var)) {
      pred_data$.lm_group <- factor(pred_data[[group_var]], levels = levels(data$.lm_group))
    }

    group_inputs <- pred_data[, combo_cols, drop = FALSE]
    pred_data$group <- as.integer(do.call(base::interaction, c(group_inputs, drop = TRUE)))

    pred <- stats::predict(fit, newdata = pred_data, se.fit = TRUE)
    pred_data$y <- as.numeric(pred$fit)

    if (isTRUE(se)) {
      crit <- stats::qt((1 + level) / 2, df = stats::df.residual(fit))
      pred_data$se <- as.numeric(pred$se.fit)
      pred_data$ymin <- pred_data$y - crit * pred_data$se
      pred_data$ymax <- pred_data$y + crit * pred_data$se
    }

    aesthetic_cols <- intersect(
      c("colour", "color", "fill", "alpha", "linetype", "linewidth", "size", "shape"),
      names(data)
    )

    for (aesthetic in aesthetic_cols) {
      if (!is.null(group_var) && identical(aesthetic, group_var)) {
        next
      }

      pred_data[[aesthetic]] <- vapply(
        seq_len(nrow(pred_data)),
        function(i) {
          rows <- data$PANEL == pred_data$PANEL[i] & data$x == pred_data$x[i]
          if (!is.null(group_var)) {
            rows <- rows & data[[group_var]] == pred_data[[group_var]][i]
          }
          data[[aesthetic]][which(rows)[1]]
        },
        data[[aesthetic]][1]
      )
    }

    pred_data
  }
)


stat_lm_means <- function(mapping = NULL,
                          data = NULL,
                          geom = "pointrange",
                          position = "identity",
                          ...,
                          interaction = FALSE,
                          se = TRUE,
                          level = 0.95,
                          na.rm = FALSE,
                          show.legend = NA,
                          inherit.aes = TRUE) {
  ggplot2::layer(
    data = data,
    mapping = mapping,
    stat = StatLmMeans,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      interaction = interaction,
      se = se,
      level = level,
      na.rm = na.rm,
      ...
    )
  )
}

partial_r2 <- function(model, details = FALSE) {
  if (!inherits(model, "lm")) {
    stop("`model` must be an `lm` or `honest_lm` object.", call. = FALSE)
  }

  details <- isTRUE(details)
  term_labels <- attr(stats::terms(model), "term.labels")

  if (length(term_labels) == 0) {
    stop("`model` must contain at least one model term.", call. = FALSE)
  }

  lm_model <- model
  class(lm_model) <- setdiff(class(lm_model), "honest_lm")

  drop_table <- stats::drop1(lm_model, test = "F")
  drop_df <- as.data.frame(drop_table)
  drop_df$term <- row.names(drop_df)
  drop_df <- drop_df[drop_df$term != "<none>", , drop = FALSE]
  drop_df <- drop_df[!is.na(drop_df$Df) & drop_df$Df > 0, , drop = FALSE]

  ss_full_error <- stats::deviance(lm_model)
  ss_reduced_error <- drop_df$RSS
  delta_ss <- drop_df[["Sum of Sq"]]

  partial <- delta_ss / ss_reduced_error
  f2 <- delta_ss / ss_full_error

  out <- tibble::tibble(
    term = drop_df$term,
    df = drop_df$Df,
    partial_r2 = as.numeric(partial),
    f2 = as.numeric(f2)
  )

  if (details) {
    out$statistic <- as.numeric(drop_df[["F value"]])
    out$p_value <- as.numeric(drop_df[["Pr(>F)"]])
    out$ss_full_error <- rep(ss_full_error, nrow(out))
    out$ss_reduced_error <- as.numeric(ss_reduced_error)
    out$delta_ss <- as.numeric(delta_ss)
  }

  out
}

penguins <- penguins |>
  filter(!is.na(sex),!is.na(body_mass),!is.na(flipper_len))

full_model <- lm(flipper_len ~ body_mass  + species +  sex, data = penguins)

```

:::aside
.
:::


```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(emmeans)
library(parameters)
source("../../_common.R") 

```


::: {.motivation style="background-color: #ffe6f7; padding: 10px; border: 1px solid #ddd; border-radius: 5px;"}

**Motivating example:** Throughout this section I have sprinkled in R code about how to do things, but this code was mixed with explanations, theory, and the like. Now that you understand the concepts of a multiple regression, you want a quick resource about how to do this in R.  

**Learning goals: By the end of this section, you should be able to:**

- Complete a first-pass analysis of data. 

:::

---

::: warning
## This is not enough

Just copying and pasting code from this section will not be a strong analysis. Please read the entire chapter! Rather, this serves as a "first-pass" resource for a more detailed analysis.
:::

### Getting started with multiple regression

Let's first load the key packages we use to assist us in multiple regression:

```{r}
#| warning: false
#| message: false
library(car)
library(ggplot2)
library(dplyr)
library(honestlm)
library(emmeans)
library(modelbased)
library(parameters)
library(broom)
penguins <- penguins |>
  filter(!is.na(sex),!is.na(body_mass),!is.na(flipper_len))
```

```{r}
#| echo: false

# Format values using three significant figures
fmt_sig <- function(x, digits = 3) {
  vapply(
    x,
    function(z) {
      if (is.na(z)) return(NA_character_)
      if (z == 0) return("0")

      decimal_places <- max(
        digits - 1 - floor(log10(abs(z))),
        0
      )

      formatC(
        z,
        format = "f",
        digits = decimal_places
      )
    },
    character(1)
  )
}

format_means_table <- function(model, by) {

  means <- modelbased::estimate_means(
    model,
    by = by
  ) |>
    as.data.frame() |>
    dplyr::mutate(
      `95% CI` = paste0(
        "[",
        fmt_sig(CI_low), ", ",
        fmt_sig(CI_high),
        "]"
      )
    ) |>
    dplyr::select(
      -tidyselect::any_of(
        c("CI", "CI_low", "CI_high", "df_error")
      )
    ) |>
    dplyr::relocate(
      `95% CI`,
      .after = SE
    )

  # Numeric columns to format and align
  number_cols <- names(means)[
    vapply(means, is.numeric, logical(1))
  ]

  # Format p separately if the output happens to include it
  ordinary_number_cols <- setdiff(number_cols, "p")

  table <- means |>
    gt::gt() |>

    gt::cols_label(
      Mean = "Mean",
      SE = "SE",
      `95% CI` = "95% CI"
    ) |>

    gt::fmt_number(
      columns = tidyselect::all_of(ordinary_number_cols),
      n_sigfig = 3,
      use_seps = FALSE
    ) |>

    gt::cols_align_decimal(
      columns = tidyselect::all_of(ordinary_number_cols)
    ) |>

    gt::cols_align(
      align = "center",
      columns = `95% CI`
    ) |>

    gt::tab_style(
      style = gt::cell_borders(
        sides = "right",
        color = "grey80",
        weight = gt::px(1)
      ),
      locations = list(
        gt::cells_body(
          columns = tidyselect::everything()
        ),
        gt::cells_column_labels(
          columns = tidyselect::everything()
        )
      )
    )

  # Apply conventional p-value formatting only when p exists
  if ("p" %in% names(means)) {
    table <- table |>
      gt::fmt(
        columns = p,
        fns = function(x) {
          ifelse(
            x < 0.001,
            "<0.001",
            formatC(x, format = "f", digits = 3)
          )
        }
      ) |>
      gt::cols_align(
        align = "right",
        columns = p
      )
  }

  table
}
```
---

### Model building

For this example, I will take the penguins data set and model flipper length as a function of body mass, sex, and species. 

```{r}
#| column: page-right
#| output: false
full_model <- lm(flipper_len ~ body_mass  + species +  sex, data = penguins)
coef(full_model)
```

```{r}
#| echo: false
full_model |>
  coef() |>
  signif(digits = 3) |>
  t() |>
  knitr::kable()
```

From these coefficients, we can build a linear model.  

$$
\widehat{\text{flip len}_i}
=
`r round(coef(full_model)[["(Intercept)"]], 0)`
+
`r signif(coef(full_model)[["body_mass"]], 3)`
(\text{body mass}_i)
+
`r round(coef(full_model)[["speciesChinstrap"]], 2)`
(\text{Chinstrap}_i)
+
`r round(coef(full_model)[["speciesGentoo"]], 1)`
(\text{Gentoo}_i)
+
`r round(coef(full_model)[["sexmale"]], 2)`
(\text{male}_i)
$$


Remember that levels of categorical "indicator variables" (e.g. male, Gentoo, etc)  are 0 if you are not that thing, and 1 if you are. So, for example, an Adelie female has a flipper length of $`r round(coef(full_model)[["(Intercept)"]], 0)`+`r signif(coef(full_model)[["body_mass"]], 3)` \times \text{its body mass}$, while an  Adelie male has a flipper length of $`r round(coef(full_model)[["(Intercept)"]], 0)` + `r round(coef(full_model)[["sexmale"]], 2)`+`r signif(coef(full_model)[["body_mass"]], 3)` \times \text{its body mass.}$



---

### Visualization 

The next step is to look at the data. Remember that 

1. There are numerous ways to visualize multivariate data.   
2. Slopes and means from [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) and [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html), respectively, are specific to each group and do not necessarily correspond to predictions from a simple additive model. 

So, rather than making you a static plot, I am providing this webR environment, with my [`honestlm`](https://yanivjb.github.io/honestlm/index.html) package loaded, with [`geom_lm_smooth()`](https://yanivjb.github.io/honestlm/reference/geom_lm_smooth.html) and [`stat_lm_means()`](https://yanivjb.github.io/honestlm/reference/stat_lm_means.html) ready to go! 

Have a look at the output and see what you notice. Then try swapping mapping of sex and species (i.e. facet by sex and color by species). Then try putting body mass on the y. What do you notice? How does this guide your understanding of the data and model?


```{webr-r}
#| autorun: true
#| fig-height: 2.5
penguins                            |>
  ggplot(aes(x = body_mass, 
             y = flipper_len, 
             color = sex))          +
  facet_wrap(~species)+
  geom_point(alpha = .3)            +
  geom_lm_smooth()

penguins                           |>
  ggplot(aes(x = sex, 
             y = flipper_len, 
             color = sex))         +
  facet_wrap(~species)+
  geom_jitter(height=0, width = .1)+
  stat_lm_means(color = "black",
                geom = "errorbar", 
                width = .2) 
```

---

### Evaluating assumptions

Now we are ready to evaluate assumptions. 

- First we can evaluate the possibility of multicollinearity with the [`vif()`](https://search.r-project.org/CRAN/refmans/car/html/vif.html) function from the [`car`](https://cran.r-project.org/web/packages/car/index.html) package: 

```{r}
vif(full_model)
```

While we see some association among predictors, I don't think we're in the danger zone yet.  

- Next we evaluate assumptions of linearity, normality of residuals, and homoscedasticity with diagnostic plots:

```{r}
#| eval: false
plot(full_model)
```

```{r}
#| column: page-right
#| fig-width: 12
#| fig-height: 3
#| echo: false
par(mfrow = c(1, 4))
plot(full_model)
par(mfrow = c(1, 1))
```

These plots all seem reasonably good. There is a slight deviation from normality assumptions but this does not worry me!



---

### Parsing effect sizes

We can now compare effect sizes with the [`partial_r2()`](https://yanivjb.github.io/honestlm/reference/partial_r2.html) function in my [`honestlm`](https://yanivjb.github.io/honestlm/index.html) package. 

```{r}
partial_r2(full_model)
```

We see that compared to species and body mass, sex explains a relatively small proportion of individual variation in flipper length, after adjusting for all associations among explanatory variables. 

--- 


### Evaluating significance 


Making sure to use car's [`Anova()`](https://search.r-project.org/CRAN/refmans/car/html/Anova.html) function, we find strongly significant association between all explanatory variables and flipper length, even after adjusting for associations among explanatory variables. 

```{r}
Anova(full_model, type = "II")
```

With a post-hoc test, we find that the flipper length of all species differ from one another after adjusting for associations between explanatory variables.

```{r}
emmeans(full_model, ~ species) |>
  contrast(method = "pairwise", adjust = "tukey")
```


--- 


### Summarizing uncertainty

Here, I will only use the [`model_based`](https://easystats.github.io/modelbased/) and the related [`parameters`](https://easystats.github.io/parameters/) (both in the [`easystats`](https://easystats.github.io/easystats/) family of packages) packages to summarize uncertainty. This deviation from the text just shows you other ways to do things! Pick your favorite.

- Uncertainty in model coefficients:

```{r}
#| eval: false
model_parameters(full_model)
```


```{r}
#| echo: false

# Format values using three significant figures
fmt_sig <- function(x, digits = 3) {
  vapply(
    x,
    function(z) {
      if (is.na(z)) return(NA_character_)
      if (z == 0) return("0")

      decimal_places <- max(
        digits - 1 - floor(log10(abs(z))),
        0
      )

      formatC(
        z,
        format = "f",
        digits = decimal_places
      )
    },
    character(1)
  )
}

params <- model_parameters(full_model) |>
  as.data.frame() |>
  dplyr::mutate(
    # Preserve literal square brackets in the rendered table
    Parameter = gsub(
      "[",
      "&#91;",
      Parameter,
      fixed = TRUE
    ),
    Parameter = gsub(
      "]",
      "&#93;",
      Parameter,
      fixed = TRUE
    ),

    `95% CI` = paste0(
      "[",
      fmt_sig(CI_low), ", ",
      fmt_sig(CI_high),
      "]"
    )
  ) |>
  dplyr::rename(
    t_328 = t
  ) |>
  dplyr::select(
    -CI,
    -CI_low,
    -CI_high,
    -df_error
  ) |>
  dplyr::relocate(
    `95% CI`,
    .after = SE
  )

# Numeric columns to format and align,
# excluding p because it receives special formatting
number_cols <- names(params)[
  vapply(params, is.numeric, logical(1))
]

number_cols <- setdiff(number_cols, "p")

params |>
  gt::gt() |>

  # Render the HTML entities as literal square brackets
  gt::fmt_markdown(
    columns = Parameter
  ) |>

  gt::cols_label(
    Parameter = "Parameter",
    Coefficient = "Coefficient",
    SE = "SE",
    `95% CI` = "95% CI",
    t_328 = gt::md("$t_{328}$"),
    p = gt::md("$p$")
  ) |>

  # Three significant figures
  gt::fmt_number(
    columns = tidyselect::all_of(number_cols),
    n_sigfig = 3,
    use_seps = FALSE
  ) |>

  # Display very small p-values as <0.001
  gt::fmt(
    columns = p,
    fns = function(x) {
      ifelse(
        x < 0.001,
        "<0.001",
        formatC(x, format = "f", digits = 3)
      )
    }
  ) |>

  # Align ordinary numeric columns at decimal points
  gt::cols_align_decimal(
    columns = tidyselect::all_of(number_cols)
  ) |>

  # Center confidence intervals
  gt::cols_align(
    align = "center",
    columns = `95% CI`
  ) |>

  gt::cols_align(
    align = "right",
    columns = p
  ) |>

  # Add vertical lines between columns
  gt::tab_style(
    style = gt::cell_borders(
      sides = "right",
      color = "grey80",
      weight = gt::px(1)
    ),
    locations = list(
      gt::cells_body(
        columns = tidyselect::everything()
      ),
      gt::cells_column_labels(
        columns = tidyselect::everything()
      )
    )
  )
```


::: warning
## Ignore these p-values
Remember, the p-values here refer to the deviation of model coefficients from a null value of zero. This is rarely what we care about. I suggest ignoring them. For term-level tests, use the Type II sums of squares approach above.
:::

- Uncertainty in adjusted means:


```{r}
#| eval: false
estimate_means(full_model, by="sex")
```

```{r}
#| echo: false

format_means_table(
  full_model,
  by = "sex"
)
```


```{r}
#| eval: false
estimate_means(full_model, by="species")
```

```{r}
#| echo: false

format_means_table(
  full_model,
  by = "species"
)
```