## • 3. Reproducibility summary {.unnumbered}

**Links to:** [Summary](#reproducible_summarySummary),  [Chatbot Tutor](#reproducible_summaryChatbot), [Practice Questions](#reproducible_summaryPractice_questions),  [Glossary](#reproducible_summaryGlossary_of_terms), [R functions](#reproducible_summaryNew_r_functions) and [R packages](#reproducible_summaryR_packages_introduced), and [Additional resources](#reproducible_summaryAdditional_resources).    


```{r}
#| echo: false
#| message: false
#| warning: false
library(knitr)
library(dplyr)
library(readr)
library(webexercises)
library(tidyr)
library(ggplot2)
library(readr)
library(dplyr)
#https://coursekata.org/preview/ec26c7bd-0ade-480b-a307-13c141fe28cb/program?version=76314360-dd94-4593-bf1c-4ce440b73eaa&book=3cc54c0e-3b8a-4804-9a2e-186c8c2bf28f
```

## Chapter Summary  {#reproducible_summarySummary}  

Science is hard. But, you can make it less painful by taking good care of your data, and saving your code which you make easy-to-understand and re-run.



### Chatbot tutor  {#reproducible_summaryChatbot}

:::tutor
Please interact with this custom chatbot ([**ChatGPT link here**](https://chatgpt.com/g/g-699602e563d8819187e3ce112c2466b5-reproducible-research-2026), [**Gemini link here**](https://gemini.google.com/gem/1bzp27nr6_xCgTN9cdjf4S1aBH1TLyhLm?usp=sharing)) I have made to help you with this chapter. I suggest interacting with at least ten back-and-forths to ramp up  and then stopping when you feel like you got what you needed from it. 
:::


### Practice Questions {#reproducible_summaryPractice_questions}


Try these questions!   



:::exercises



**Q1)** What is the biggest mistake in the table below?  `r longmcq(c("ID should be lower case","Its perfect, change nothing", answer = "the column name, *weight* is not sufficiently descriptive, it should include the units.", "date_collected_empty_means_same_as_above is too wordy, replace with *date*", answer = "Values for date_collected_empty_means_same_as_above are implied.","date is in Year-Month-Day format, while Month-Day-Year format is preffered."))`  

```{r}
#| echo: false
#| message: false
#| warning: false
library(lubridate)
tibble(
  ID             = c("1-A1","1-1B","3-7","2-B","1-A5"),
  weight         = c(104, 210, 150, 176,110),
  date_collected_empty_means_same_as_above = c("2024-03-01", "","" , "2024-03-15","" ))|>
  kable()
```
`r hide("Click here for explanation")`

While some of these (like the long name for date) are clearly shortcomings, spreadsheets should never leave values implied. 

`r unhide()`. 

---  


**Q2)** What would you expect in a data dictionary accompanying the table above? (select all correct) `r longmcq(c(answer = "The units for weight.",answer = "A statement that date is in Year-Month-Day format", answer = "A statement explaining that in the  date collected column, empty means same as above."))`  


---  


**Q3)** How do you read data from a Excel sheet, called *raw_data* in an Excel filed named *bird_data.xlsx* located  inside the R project  you are working in? `r longmcq(c( "You cannot load excel files into R. You must save it as a csv, and read it in with read_csv().",answer = "Assuming the readxl package is installed and loaded, type read_xlsx(file = \"bird_data.xlsx\", sheet = \"raw_data\").",  "While you can read excel into R, you cannot specify the sheet."))`  

---  

**Q4)** What should you do to make code reproducible?  (pick the best answer) `r longmcq(c("Specify the working directory with setwd()","Show the packages installed with install.packages()", answer = "Restart R once your done, and rerun your script to see if it works"))`

---


**Q5)** As we saw in this chapter, R has a built-in dataset called *iris*. You can look at it or give it to functions by typing *iris*. Which variable type is the *Species* in the *iris* dataset? `r longmcq(c("numeric", "logical", "character", answer = "factor"))` 

---

**Q6)** Consider the plot generated by the [code in the previous section](#reproducible-code-example). The plot consists of "small multiples" (or in ggplot language "facets"). The facet on the far right is **pink**. What is the facet on the far right?  `r fitb(answer = "NA")`

:::

**For the following question consider the** ***diabetes dataset*** available at: https://raw.githubusercontent.com/ybrandvain/datasets/refs/heads/master/diabetes.csv 

:::exercises

**Q7)** What is the first column in the diabetes dataset that will change its name after going through the clean_names() function?: `r fitb(answer = "stab.glu")`


:::

---

For the next questions, consider the R script below, which is (sadly) not far from bad code I often have to handle as a data editor.


```{r}
#| eval: false
# My R script

setwd("~/ybrandva/Desktop/silphium_project")

#### load data 
silphium_data <- read_csv("mydata.csv")
View(silphium_data)


ggplot(silphium_data, aes(x = location, y = yield))+
  geom_point()
```

:::exercises
**Q8)** The script above worked fine on my computer yesterday, but today I opened up a new R session on the very same computer and it failed. What went wrong? `r mcq(c("I set my working directory", "I used an absolute path, not a relative path",answer = "I did not load the libraries in my script", "I used the View function"))`

---

**Q9)** what about the script would make it work on my computer (once the thing above is fixed) but not on another computer? `r mcq(c(answer = "I set my working directory", "I used an absolute path, not a relative path", "I did not load the libraries in my script", "I used the View function"))`

---

**Q10)** What about the script is annoying for someone using this code (once it works) and should be removed, but doesnt stop our code from working? `r mcq(c( "I set my working directory", "I used an absolute path, not a relative path", "I did not load the libraries in my script",answer = "I used the View function"))`
::: 


----



### Glossary of Terms {#reproducible_summaryGlossary_of_terms} 

:::glossary


**Absolute Path** – A file location specified from the root directory (e.g., `/Users/username/Documents/data.csv`), which can cause issues when sharing code across different computers. Using **relative paths** instead is recommended.  

**Data Dictionary** – A structured document that defines each variable in a dataset, including its name, description, units, and expected values. It helps ensure data clarity and consistency.  

**Data Validation** – A method for reducing errors in data entry by restricting input values (e.g., dropdown lists for categorical variables, ranges for numerical values).  

**Field Sheet** – A structured data collection form used in the field or lab, designed for clarity and ease of data entry.  

**Metadata** – Additional information describing a dataset, such as when, where, and how data were collected, the units of measurement, and details about the variables.  

**R Project** – A self-contained environment in RStudio that organizes files, code, and data in a structured way, making analysis more reproducible.  

**Raw Data** – The original, unmodified data collected from an experiment or survey. It should always be preserved in its original form, with any modifications performed in separate scripts.  

**README File** – A text file that provides an overview of a dataset, including project details, data sources, file descriptions, and instructions for use.  

**Reproducibility** – The ability to re-run an analysis and obtain the same results using the same data and code. This requires careful documentation, structured data storage, and clear coding practices.  

**Relative Path** – A file path that specifies a location relative to the current working directory (e.g., `data/my_file.csv`), making it easier to share and reproduce analyses.  

**Tidy Data** – A dataset format where each variable has its own column, each observation has its own row, and each value is in its own cell.  


:::

---

### Key R functions  {#reproducible_summaryNew_r_functions}  



:::functions



**📥 Data import**    

* - **[`read_csv("file.csv")`](https://readr.tidyverse.org/reference/read_delim.html)** – Reads a CSV file into R as a tibble (from the [`readr`](https://readr.tidyverse.org/) package).    
* **[`read_xlsx("file.xlsx", sheet = "sheetname")`](https://readxl.tidyverse.org/reference/read_excel.html)** – Reads an excel sheet into R as a tibble (from the [`readxl`](https://readxl.tidyverse.org/) package).   

---

**🔍 Inspecting data**   

* **[`glimpse()`](https://dplyr.tidyverse.org/reference/glimpse.html)** — shows the structure of your data from the [`dplyr`](https://dplyr.tidyverse.org) package. 
* **[`View()`](https://tibble.tidyverse.org/reference/view.html)** — open data viewer.  

---

**🧹 Cleaning & renaming**  

* **[`clean_names(data)`](https://sfirke.github.io/janitor/reference/clean_names.html)** – Standardizes column names (from the [`janitor`](https://sfirke.github.io/janitor/index.html) package). 
* **[`rename(data, new_name = old_name)`](https://dplyr.tidyverse.org/reference/rename.html)** – Renames columns in a dataset (from the [`dplyr`](https://dplyr.tidyverse.org/) package).  

---

**🔧 Data wrangling**    

* **[`select()`](https://dplyr.tidyverse.org/reference/select.html)** — choose columns (from the [`dplyr`](https://dplyr.tidyverse.org/) package).   
* **[`pivot_longer(data, cols, names_to, values_to)`](https://tidyr.tidyverse.org/reference/pivot_longer.html)** – Converts wide-format data to long format (from the [`tidyr`](https://tidyr.tidyverse.org/) package).   

:::

---



### R Packages Introduced {#reproducible_summaryR_packages_introduced} 

:::packages


- **[`readr`](https://readr.tidyverse.org/)** – Provides fast and flexible functions for reading tabular data (here we revisited [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) for CSV files).  

- **[`dplyr`](https://dplyr.tidyverse.org/)** – A grammar for data manipulation. Here we introduced the [`rename(data, new_name = old_name)`](https://dplyr.tidyverse.org/reference/rename.html) function to give columns better names.    

- **[`tidyr`](https://tidyr.tidyverse.org/)** – Helps tidy messy data. Here we introduced [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) to make wide data long.    

- **[`janitor`](https://sfirke.github.io/janitor/)** – Cleans and standardizes data, including [`clean_names()`](https://sfirke.github.io/janitor/reference/clean_names.html) for formatting column names.  


:::





### Additional resources {#reproducible_summaryAdditional_resources} 


:::learnmore 

**R Recipes:**      

- [Read a .csv](https://posit.cloud/learn/recipes/basics/ImportA1):  Learn how to read a csv into R as a tibble.       
- [Read an Excel file](https://posit.cloud/learn/recipes/basics/ImportF): Learn how to read an excel file into R as a tibble.    
- [Obey R's naming rules](https://posit.cloud/learn/recipes/basics/BasicB3A):  You want to give a valid name to an object in R.  
- [Rename columns in a table](https://posit.cloud/learn/recipes/transform/TransformC): You want to rename one or more columns in a data frame.




**Other web resources:**    

- *[Data Organization in Spreadsheets](https://www.tandfonline.com/doi/citedby/10.1080/00031305.2017.1375989?scroll=top&needAccess=true)* [@broman2018].       

- *[Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf):* [@TidyData].     

-  *[Ten Simple Rules for Reproducible Computational Research](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003285):* [@sandve2013].       

- [NYT article: For big data scientists hurdle to insights is janitor work.](https://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html)
   
- [Style guide](https://datamgmtinedresearch.com/style): Chapter 9 of *Data management in large-scale education research* by @lewis2024.  Includes sections on [general good practices](https://datamgmtinedresearch.com/style#style-practices), [file naming](https://datamgmtinedresearch.com/style#style-file), and [variable naming](https://datamgmtinedresearch.com/style#style-varname). 

- [Data Storage and security](https://datamgmtinedresearch.com/store): Chapter 13 of *Data management in large-scale education research* by @lewis2024.  

**Videos:**  

- [Data integrity](https://www.youtube.com/embed/G9LDtFMsUmY): (By [Kate Laskowski](https://laskowskilab.faculty.ucdavis.edu/people/kate-laskowski/) who was the victim of data fabrication by her collaborator (and my former roommate) Jonathan Pruitt).    

- [Tidying data with `pivor_longer`](https://www.youtube.com/embed/qivE6exNsZI?end=340) (From Stat454)




::: 





